SAT Solving for Vesicle Traffic Systems in Cells

### 2017-06-16 paper/encoding.tex

In this section, we will present our encoding of the search problem.

@@

Basic notation for the formula we are trying to generate.
\begin{enumerate}
\item edge = e(i,j,q)
      There is an qth edge between node i and j.

\item dump = d(i,j,q)
       drop edge(i,j,q) from the graph.

@@

\subsubsection{Regulation on nodes and edges}
Regulation of a model is based on the kind of the model under consideration. for example model 1 (-V 1) has two different arbitrary Boolean function dictating the activity of the molecules on edge and nodes and for variation 3 (-V 3) activity on edge is driven by pairing inhibition.

We have build constraints A0, A1 for these two regulations. \newline

A0: No regulation on nodes.
\[ \bigwedge\limits_{i,k} n_{i,k} == a_{i,k} \, \]

A0: Boolean regulation on nodes. Activity of a molecule k on a node is defined as a Boolean function of presence of other molecule present on that node.
\[ \bigwedge\limits_{i,k} n_{i,k} \supset a_{i,k} ==   f_n[k] (\bigvee_{ k' != k} n_{i,k'}) \, \]

A1: No regulation on edges.
\[ \bigwedge\limits_{i,j,q,k} e_{i,j,q,k} == b_{i,j,q,k} \, \]

A1: Boolean regulation on edges. Activity of a molecule k on a node is defined as a Boolean function of presence of other molecule present on that node.
\[ \bigwedge\limits_{i,j,q,k} e_{i,j,q,k} \supset b_{i,j,q,k} == f_e[k](\bigvee_{k' != k} e_{i,j,q,k'}) \, \]

A1: Pairing inhibition. Inhibition of the edges are driven by the pairing matrix.
\[ \bigwedge\limits_{i,j,q, k}  (e_{i,j,q,k} \land [ ( \bigvee_{k'} p_{k,k'} \land \bigwedge_{k' != k} p_{k,k'} \supset e_{i,j,q,k'})] ) \supset \neg b_{i,j,q,k}) \, \]

@@

\subsubsection{Constraints on the structure of the edges, nodes and pairing matrix}

@@

\subsubsection{Constraints on presence and activity of the molecule}
Few additional constrints are required on the molesules presence and activity. \newline

@@

\subsubsection{Constraints for stability condition}
The whole cell should be in stable condition i.e global flow of the molecule is balanced at every point. Each molecule leaving the node on a vesicle should come back to its source node in a cycle, i.e., \textbf{for every molecule leaving the node there exists a cycle with that molecule present on \textit{each of the edges and nodes} of the path taken}.

\bf{Reachability definition and stability condition}
We have encoded this requiremnt by using reachability definition. 
If molecule is active on an edge then it should be present on the edge. \newline

F3: stability condition. Source node is reachable by target node with that molecule present in p steps.

@@

F2: Reachability definition.
\[ \bigwedge\limits_{i,j,k,p} r_{i,j,k,p} \supset (\bigvee_{q} \, e_{i,j,q,k} \lor \bigvee_{i\neq i^{\prime}} ( \, \bigvee_{q} e_{i,i^{\prime},q,k}) \land r_{i^{\prime},j,k,p-1} ) \, \]

@@

\begin{enumerate}
\item  A \textbf{Pairing mechanism} which determines compatible Q-R pairs on vesicles and compartments that can cause fusion.
\item \textbf{Regulatory mechanisms} on edges and on nodes (possibly distinct) which regulate molecules activity based on the presence/absence of other molecules on the corresponding node or edge.
\end{enumerate}

F4: For an edge to be valid, at least one SNARE pair on the vesicle and target compartment must be active, and have a non-zero entry in the pairing matrix.

@@

\subsubsection{Graph connectivity constraints}
To check the whether that n connected is necessary condition, we remove (drop) n edges from the graph and if it disconnects the graph and we get an assignment we have a n connected satisfying graph. We can go down or up using -C \_ option.  \newline

@@

D4: New reachability definition for grph connectedness: dReachable. Node i,j are reachable either if there is a direct edge and its not dropped. 
Or there is an node i' such that, there is a direct edge between i,i' which is not dropped and i' and j is dReachable.

@@

>>>>

In this section, we will present our encoding of the search problem. Every variable is a Boolean bit, for example, n(0,1) represents node 0 with type 1 molecule present. Assignment to the Boolean variable n(0,1) to True will mean that at node 0 type 1 molecule is present.

@@

Basic notation for the formula we are trying to generate.
\begin{enumerate}

\item edge = e(i,j,q)      
      we will assign e(i,j,q) for a qth edge between node i and j.

\item dump = d(i,j,q)
      to represent drop of edge(i,j,q) from the graph we will use d(i,j,q).

@@

\subsubsection{Regulation on nodes and edges}
Regulation of the molecules is based on the kind of the model under consideration. For example, model 1 (-V 1) has two different arbitrary Boolean functions dictating the activity of the molecules on edge and nodes. For variation 3 (-V 3) activity on edge is driven by pairing inhibition.

We use two Boolean variables A0 and A1 to build these two constraints. \newline

A0: No regulation on nodes i.e all the present molecules on the nodes are active.
\[ \bigwedge\limits_{i,k} n_{i,k} == a_{i,k} \, \]

A0: Boolean regulation on nodes. Activity of a molecule k on the node is defined as a Boolean function of presence of other molecule present on that node.
\[ \bigwedge\limits_{i,k} n_{i,k} \supset a_{i,k} ==   f_n[k] (\bigvee_{ k' != k} n_{i,k'}) \, \]

A1: No regulation on edges i.e all the present molecules on the edges are active.
\[ \bigwedge\limits_{i,j,q,k} e_{i,j,q,k} == b_{i,j,q,k} \, \]

A1: Boolean regulation on edges. Activity of a molecule k on the node is defined as a Boolean function of presence of other molecule present on that node.
\[ \bigwedge\limits_{i,j,q,k} e_{i,j,q,k} \supset b_{i,j,q,k} == f_e[k](\bigvee_{k' != k} e_{i,j,q,k'}) \, \]

A1: Pairing inhibition. The activity of the molecules is defined by inhibition by other molecules based on the pairing matrix. 
\[ \bigwedge\limits_{i,j,q, k}  (e_{i,j,q,k} \land [ ( \bigvee_{k'} p_{k,k'} \land \bigwedge_{k' != k} p_{k,k'} \supset e_{i,j,q,k'})] ) \supset \neg b_{i,j,q,k}) \, \]

@@

\subsubsection{Constraints on the structure of the edges, nodes and pairing matrix}
A few additional constraints are required on the nodes, edges and pairing matrix to set up the essential conditions. We use Boolean variables C0-C5 to build all these constraints.

@@

\subsubsection{Constraints on presence and activity of the molecule}
Few additional constraints that are required for the molecules presence and activity is build using F0 and F1. \newline

@@

\subsubsection{Constraints for stability condition}
The whole cell should be in stable condition i.e flow of the molecule is balanced at every point and globally. Each molecule leaving its source node on a vesicle should come back in a cycle. Boolean variable F3 and F2 are used to model these constraints.

\textbf {Reachability definition and stability condition.}
We have encoded stable condition by using reachability definition. \newline

F3: Stability condition. For every leaving molecule, the source node is reachable from the target node with that molecule present, in maximum p steps.

@@

F2: Reachability definition. Node j is reachable from node i with kth molecule present in maximum p steps if either there is a direct edge between i-j with that molecule present or there is a direct edge between $i^{\prime \prime}$ $(i != i^{\prime \prime})$ and j with kth molecule present on that edge and $i^{\prime \prime}$ are reachable from i in p steps.

\[ \bigwedge\limits_{i,j,k,p} r_{i,j,k,p} \supset (\bigvee_{q} \, e_{i,j,q,k} \lor \bigvee_{i\neq i^{\prime}} ( \, \bigvee_{q} e_{i,i^{\prime},q,k}) \land r_{i^{\prime},j,k,p-1} ) \, \]

@@

\begin{enumerate}
\item  A \textbf{pairing mechanism} which determines compatible Q-R pairs on vesicles and compartments that can cause fusion.
\item \textbf{regulatory mechanisms} on the edges and nodes (possibly distinct) which regulates molecules activity based on the presence/absence of other molecules on the corresponding node or edge.
\end{enumerate}

Boolean variable F4 and F5 are used to model these constraints.

F4: For an edge to be valid, at least one SNARE pair on the vesicle and target compartment must be active and have a non-zero entry in the pairing matrix.

@@

\subsubsection{Graph connectivity constraints}
To check the whether that n connected is a necessary condition, we remove (drop) n edges from the graph and if it disconnects the graph and we get an assignment we have an n connected satisfying graph. We can go down or up using -C \_ option.  \newline

@@

D4: New reachability definition for graph connectedness: dReachable. Node i,j are reachable either if there is a direct edge and its not dropped or there is an node $i^{\prime}$ such that, there is a direct edge between i, $i^{\prime}$ which is not dropped and $i^{\prime}$ and j is dReachable.

@@

\subsection{Creating a solver and solving these constraints}
We will create a solver and add these build formula using these constraints to that and then check whether there a satisfiable assignment that satisfies all these constraints.

\begin{figure}[ht]
%\vspace{-3mm}
\begin{verbatim}
# Create a solver
s = Solver()
# Add these formulas to it.
s.add(A0, A1, C0, C1, C2, C4, C5, F0, F1, F2, F3, F4, F5, D0, D1, D2, D3)

# Check for the satisfiability assignment.
print s.check()   
\end{verbatim}
\caption{Solving for these constraints.}
\label{code:motivate}
\end{figure}

### 2017-06-16 paper/experiments.tex

\begin{enumerate}[label=\Alph*]
\item Every present molecule is considered to be active.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules. 
\item Activity of molecules on the edges is based on Boolean function of presence of other molecules
\item Activity of molecules both on the edges and nodes is based on Boolean function of presence of other molecules.
\item Activity of molecules on the edges is driven by pairing inhibition.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition.
\item Version F with additional constraint that every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition with every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other moleculess and on edge by pairing inhibition with self-edges are allowed.
\item Version F but self-edges are allowed.
\end{enumerate}

@@

\input{tbl-qf-graph}
In table~\ref{tbl:qf-results} 
\b PARTIAL ...

@@

\todo{Make table for all sizes and variants of the tool}

>>>>

\begin{enumerate}[label=\Alph*]
\item Every present molecule is considered to be active.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules. 
\item Activity of molecules on the edges is based on Boolean function of presence of other molecules
\item Activity of molecules both on the edges and nodes is based on Boolean function of presence of other molecules.
\item Activity of molecules on the edges is driven by pairing inhibition.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition.
\item Version F with additional constraint that every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition with every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other moleculess and on edge by pairing inhibition with self-edges are allowed.
\item Version F but self-edges are allowed.
\end{enumerate}
%
Our tool allows user to choose a model and the size of
the network.
%
It builds the needed constraints and applies $\zthree$
on the constraints.
%
Our tool also translates the model found by $\zthree$
into a vesicle traffic system and presents a visual
output to the user.

@@

\input{tbl-qf-graph}

In table~\ref{tab:qf-graph}, we present the running times
for the search of vesicle traffic systems of sizes 2 to 10
that satisfies the model variants.
%
The experiments were done on a machine with ... processor 
and ... RAM.
%

%
$\zthree$ was able to solve the constraints up to 10
nodes.
%

In table~\ref{tbl:qf-results} 
\b PARTIAL ...

@@
